---
title: "Mortality Exercises 3: Person Years Remaining"
author: Joshua R. Goldstein
output: html_notebook
---

In these exercises, we will calculate person-years lost from a sudden mortality crisis like Covid. Our approach will be to do the full life-table calculation and then compare this to the approximation involving $H/A$ that we did in lecture.

More future reference (but not now), see Goldstein and Lee (2020) (https://www.pnas.org/content/117/36/22035.full)





## A. Person years remaining in stationary pop with US life table

Here we're going to sum up all of the person-years in a simulated but not-totally-unrealistic population

Notes:
* We are using 2019 period life table from Human Mortality Database
* Remaining person years is a cohort concept, but we are using
 period life expectancy for simplicity here.
* We are using stationary age structure, not the observed
* To give us numbers that are the right order of magnitude, we're
 going to assume 3.5 million births.
* We're going to be a bit sloppy with distinction between l(x) vs L(x). For actual research, please be more careful.


Some preliminaries

```{r}

library(data.table)
dt <- fread("exercise_2_data.csv") ## data
source("exercise_2_funs.R") ## function
head(dt) ## just to see what's there
hx = dt$hx ## assign baseline mortatity (from 2019 period)

## to get the stationary population, we compute the life table from these mortality rates
lt = get.lt(hx) ## function to make life table, print(get.lt) for gory details
lx = lt$lx
ex = lt$ex
Lx = lt$Lx
x = lt$x
## you can try plotting a function if you want: e.g. plot(x, ex)

## To make more realistic let's have the stationary population has 3.5
## million births per year, pretty close to USA

Nx.stationary = lx * 3.5
sum(Nx.stationary) ## [1] 278.8149
```



Now we're ready to calculate remaining person years, which will be
the number of people in each age group, multiplied by their life
expectancy, summed over all ages.

```{r}
remaining.py = sum(Nx.stationary * ex)
## [1] 11606.68, or 11.6 billion
```

> Q.1 How many person years remain per person?  

> (Fun fact: in stationary population average life remaining is the same as average age of  pop! 
> Life left = life lived)


## B. Person Years Lost

Here we assume there is a "flash" of mortality, with intensity 1/3 of
all-cause annual mortality. How many person years will be lost?

```{r}
## Number of deaths at each age
Dx.flash = Nx.stationary * hx/3

lost.py = sum(Dx.flash * ex)
print(lost.py)
## [1] 14.83438, 14.8 _million_ not _billion_
```

What fraction of the original person-years of the population were lost?
```{r}
lost.py.frac = lost.py / remaining.py
print(lost.py.frac)
## [1] 0.001278089
## or
## 1/lost.py.frac
## [1] 782.4179
## a bit more than 1/800
```


> Q2. How many person years remain _per person_ after the flash of
> mortality? How many years are lost _per person_? Is this a few
>  years, months, weeks, days, or hours, ...?


## C. Share of person-years lost (using our approximation)

Compare your answer to our approximation which tells us that the
fraction of person years lost in this case should be about H * delta/A

We leave it to you to calculate $H$ and $A$ that correspond to the stationary population to the 2019 US life table

(Note: The values should be something close to $H = 0.15$ and $A = 40$. If you're having trouble estimating the exact values, you can use these in order to answer the Qs)

```{r}
delta = 1/3
A = NA ## fill in a formula for average age of stationary population. 
## Hint: fraction aged x is Lx / sum(Lx)

H = NA  ## Fill in formula for Keyfitz's entropy H. 
## Hint: Take a look at exercise 2.

lost.py.frac.hat  = H * delta / A
print(lost.py.frac.hat)
```
> Q3. How did we do? Are we close to life-table based calculation?

> Q4. On twitter, one of the workshop participants didn't like this
> approach, saying that the tiny number was misleading --  and said that
> it was like comparing the economic damage done by Hurricane Katrina
> in New Orleans to the GDP of the entire USA. It would seem
> small, but is a minimizing way to think about the actual damage of
> the Hurricane. What do _you_ think?

Note: When comparing past mortality crisis, Goldstein and Lee had similar concerns and ended up divided remaining-years-lost from each crisis by the remaining-years-lost in a corresponding non-crisis year. You can look at their Fig. 4 to see the results of this analysis. (https://www.pnas.org/content/117/36/22035.full)

## Congratulations!

You've finished the last exercise of the day!




