---
title: "Mortality Exercises 2: Entropy"
author: Joshua R. Goldstein
output: html_notebook
---

In these exercises, we build on what you did in the pre-lab exercises, using the "entropy" approximation, and applying the logic to Covid.

Note 1: There were about half-a-million additional Covid deaths in the US
in 2020, compared to 3 million that would have been expected
without Covid. So the increase in death rates was about 1/6 and we're going to assume $\delta = 1/6$.

Note 2: We're going to use the following naming conventions.

* suffix of ".lt" means it's  based on the calculation of the full life table, assuming proportional change in hazards
* suffix of  ".entropy" means it's based on entropy approximation
* suffix of ".realistic" means it's based on full calculation of life table assuming a realistic, not-exactly proportional increase in hazards at all ages.

## Preliminaries

```{r}
## Input data and convenience functions
library(data.table)
source("exercise_2_funs.R") ## get.lt() and get.e0()
dt <- fread("exercise_2_data.csv") ## contains baseline hazards 

## Create variables from data
x = dt$x ## age
hx.base = dt$hx ## US all-cause mortality by age in 2019 (without covid)
```



##   A. Use full life table to calculate change e0, assuming proportionality


In order to calculate life expectancy without Covid, we use age-specific mortality rates (hazards) fro 2019 US and our function that calculates life expectancy at birth from hazards. Note: If you're curious about what get.e0() does, just type print(get.e0).

```{r}
e0.base = get.e0(hx.base) ## [1] 79.11594
print("e0.base")
print(e0.base)
```

Now calculate e0.covid.lt, assuming covid is proportional to
baseline mortality and using our life-table function to calculate
life expectancy from hazards
```{r}
delta = 1/6 ## increase in hazards at all ages from Covid
e0.covid.lt = get.e0(hx.base * (1 + delta)) ## recalculation with new hazards
print("e0 with covid, life table calculation using proportional covid")
print(e0.covid.lt)
```


And let's also calculate drop in e0.

```{r}
## By how many years does e0 change?
e0.change.lt = e0.covid.lt - e0.base
print("e0 change, life table calculation using proportional covid")
print(e0.change.lt)
```

> Q1. Did life expectancy decline by 1/6th? If not, how much did it 
> decline by, and what value of entropy does this suggest?



##   B. Compare to entropy result 

We start by getting Lx using our built in functions. Note: we could also calculate sum(lx *log(lx))/sum(lx), but better to use Lx.

```{r}
lt = get.lt(hx.base) ## using our
x = lt$x
Lx = lt$Lx
```

Now compute Keyfitz $H$
```{r}
e0 = sum(Lx)
keyfitz.H = -sum(log(Lx)*Lx)/e0 
print(keyfitz.H)

## Note: you can also calculate entropy as sum(ex*Lx*hx)/e0. If we had
## even finer age groups we would get the same number. With the 1-year
## age groups we get very close.
```

> Q2. What is your estimate of H? Is this close to what you expected in Q1?

Now we can use $H$ to calculate effect on life expectancy

```{r}
## calculate estimate of proportional decline
e0.proportional.change.entropy = -keyfitz.H * delta

## convert to estimate of absolute decline
e0.change.entropy = e0.base * e0.proportional.change.entropy
print(e0.change.entropy)
```

Now let's compare:
```{r}
result = c(e0.change.lt = e0.change.lt,
           e0.change.entropy= e0.change.entropy)
print(round(result, 1))
```

> Q3. Does the entropy approximation give exactly the same answer as
> the life table calculation? Any ideas why they might differ?

## C. Using a more realistic schedule

Instead of using proportional assumption, we can use observed age 
pattern of Covid mortality. For this we will use the "standard" 
estimated by Goldstein and Lee early in the pandemic.


```{r}
hx.covid.realistic = dt$hx.covid ## Covid mortality in US
## Note: hx.covid.realistics is based on standard age schedule from
## Goldstein and Lee, rescaled to produce exactly the same number of
## deaths as a 1/6 increase in mortality at all ages.
```

And now use this hazard schedule to recalculate effect on e0
```{r}
e0.covid.realistic = get.e0(hx.base + hx.covid.realistic)

## estimated drop
e0.change.realistic = e0.covid.realistic - e0.base

## print out results
result.again = c(e0.change.lt = e0.change.lt,
                 e0.change.entropy= e0.change.entropy,
                 e0.change.realistic = e0.change.realistic)
print(round(result.again, 1))
```

> Q4. You should see that life expectancy decline is smaller using
> the more realistic schedule. Why do you think this might be? Answer
> before (and after) looking at the plot below.

Here's a plot to compare the age-schedules

```{r}


hx.covid.prop = delta * hx.base ## Covid mortality in US, assuming proportional increase
par(mfrow = c(2,2))
plot(x, hx.covid.realistic, type = "l", col = "blue", lty = 1, log = '',
     main = "Covid mortality estimates")
lines(x, hx.covid.prop, col = "red", lty=2)
legend("topleft",
       legend = c("Proportional",
                  "'Realistic'"),
       col = c("red", "blue"),
       lty = c(2,1),
       bty = "n")
## log
plot(x, hx.covid.realistic, type = "l", col = "blue", lty = 1, log = 'y',
     main = "Covid mortality estimates (log)")
lines(x, hx.covid.prop, lty = 2, col = "red")
legend("topleft",
       legend = c("Proportional",
                  "'Realistic'"),
       col = c("red", "blue"),
       lty = c(2,1),
       bty = "n")
## Note: the blue line looks perfectly straight but in actuality it's slope does vary a bit
## zoom in on at slopes of log plot
plot(x[-1], diff(log(hx.covid.realistic)),
     main = "Slopes of log of \n  realistic schedule")

```

You should see that observed Covid mortality increases by a bit more than proportionately at older ages and a bit less at younger ages. (Although there are many data issues that mean we're still not that certain about the exact Covid mortality rates.)

You can also double check that the two schedules correspond to the same number of deaths.

```{r}
Nx.usa <- dt$Nx
print(sum(hx.covid.prop*Nx.usa))
print(sum(hx.covid.realistic*Nx.usa))
## the should match!
```

## Congratulations. 

You've finished the 2nd set of exercises of Day 2 of the workshop. Only one more to go!


